---
title: "NYPD Shooting Incident Analysis"
author: "Anastasiia Stukova"
date: "2025-01-21"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(scales)
options(repos = c(CRAN = "https://cloud.r-project.org"))

```
## <span style="color: darkblue; font-family: Arial; text-align: center;">Executive Summary</span>

This report presents an in-depth analysis of shooting incidents in New York City based on publicly available NYPD data. We aim to understand patterns, identify high-risk areas and times, and provide data-driven recommendations for improving public safety.


### Methodology
Our analysis combines statistical analysis with data visualization to examine:
- Time-based patterns (annual trends, daily patterns)
- Geographic distribution
- Demographic profiles
- Fatality rates and risk factors


## <span style="color: darkblue; font-family: Arial; text-align: center;">Discussion</span>

Our analysis revealed several critical insights:

1. **Temporal Patterns**
  - Clear peak hours during late night (10 PM - 2 AM)
  - Seasonal variations with summer spikes
  - Notable increase post-2020

2. **Geographic Distribution**
  - Brooklyn shows highest concentration
  - Significant borough-level disparities
  - Specific high-risk neighborhoods identified

3. **Demographic Impact**
  - Males 18-44 most affected
  - Young adults at particular risk
  - Notable fatality rate variations

## <span style="color: darkblue; font-family: Arial; text-align: center;">Impact and Implications</span>

These findings have significant implications for:
- Law enforcement resource allocation
- Community safety programs
- Public policy development
- Emergency response planning



## <span style="color: darkblue; font-family: Arial; text-align: center;">Data Preparation</span>
```{r}



csv_url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

shooting_data <- read_csv(csv_url) %>%
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    OCCUR_TIME = hms(OCCUR_TIME),
    year = year(OCCUR_DATE),
    month = month(OCCUR_DATE),
    hour = hour(OCCUR_TIME)
  )

# Summary of the dataset
glimpse(shooting_data)


```
## <span style="color: darkblue; font-family: Arial; text-align: center;">Trends</span>
```{r}

```
<span style="color: darkblue; font-family: Arial; text-align: center;">Annual Trends</span>
```{r}

annual_trend <- shooting_data %>%
  group_by(year) %>%
  summarise(
    incidents = n(),
    fatalities = sum(STATISTICAL_MURDER_FLAG == "true", na.rm = TRUE)
  )

ggplot(annual_trend, aes(x = year)) +
  geom_line(aes(y = incidents, color = "Incidents"), size = 1.2) +
  theme_minimal() +
  labs(title = "Annual Shooting Incidents",
       y = "Count") +
  theme(legend.position = "bottom")




```
<span style="color: darkblue; font-family: Arial; text-align: center;">Hourly Distribution</span>
```{r}



hourly_dist <- shooting_data %>%
  group_by(hour) %>%
  summarise(incidents = n())

ggplot(hourly_dist, aes(x = hour, y = incidents)) +
  geom_bar(stat = "identity", fill = "#2c7fb8") +
  theme_minimal() +
  labs(
    title = "Hourly Distribution of Shooting Incidents",
    x = "Hour of Day",
    y = "Number of Incidents"
  ) +
  scale_x_continuous(breaks = 0:23)


```
<span style="color: darkblue; font-family: Arial; text-align: center;">Borough Analysis</span>
```{r}

borough_analysis <- shooting_data %>%
  group_by(BORO) %>%
  summarise(
    total_incidents = n(),
    fatal_incidents = sum(STATISTICAL_MURDER_FLAG == "true", na.rm = TRUE),
    fatality_rate = round((fatal_incidents / total_incidents) * 100, 2)
  )

ggplot(borough_analysis, aes(x = reorder(BORO, -total_incidents), y = total_incidents)) +
  geom_bar(stat = "identity", fill = "#31a354") +
  theme_minimal() +
  labs(
    title = "Shooting Incidents by Borough",
    x = "Borough",
    y = "Number of Incidents"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
<span style="color: darkblue; font-family: Arial; text-align: center;">Key Statistics</span>
```{r}

# Creating a table with key statistics
stats_summary <- data.frame(
  Metric = c(
    "Total Incidents",  # Total number of incidents
    "Total Fatalities", # Total number of fatalities
    "Most Affected Borough", # Borough with the highest number of incidents
    "Peak Hour" # Hour with the most incidents
  ),
  Value = c(
    nrow(shooting_data), # Calculates the total number of rows (incidents) in the dataset
    sum(as.logical(shooting_data$STATISTICAL_MURDER_FLAG), na.rm = TRUE), # Counts the total number of fatalities
    borough_analysis$BORO[which.max(borough_analysis$total_incidents)], # Finds the borough with the maximum incidents
    hourly_dist$hour[which.max(hourly_dist$incidents)] # Identifies the hour with the highest number of incidents
  )
)

# Displaying the key statistics table
kable(stats_summary, caption = "Key Statistics") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), # Adds styling options: striped rows, hover effect, compact layout
    full_width = FALSE # Ensures the table is not stretched to full width
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "darkblue") %>% # Styles the header row: bold text, white font, dark blue background
  column_spec(1, bold = TRUE) %>% # Highlights the first column with bold text
  column_spec(2, color = "darkgreen") # Changes the color of the values in the second column to dark green



```
<span style="color: darkblue; font-family: Arial; text-align: center;">Demographic Analysis</span>
```{r}



demographic_profile <- shooting_data %>%
  group_by(VIC_AGE_GROUP, VIC_SEX) %>%
  summarise(
    incidents = n(),
    fatalities = sum(as.logical(STATISTICAL_MURDER_FLAG), na.rm = TRUE)
  ) %>%
  arrange(desc(incidents)) %>%
  mutate(fatality_rate = round((fatalities / incidents) * 100, 2)) # 


kable(head(demographic_profile, 10), caption = "Top 10 Demographic Groups by Incidents") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Metrics" = 3)) 



```
## <span style="color: darkblue; font-family: Arial; text-align: center;">Conclusions and Recommendations</span>
```{r}



```

**Key Findings**

### **Temporal Patterns**
- **Peak incidents** occur during late-night hours, specifically around **11 PM**.
- **Seasonal trends** suggest spikes in incidents during **warmer months**.

### **Geographic Distribution**
- **Brooklyn** accounts for the highest number of incidents, with a **significant fatality rate**.
- Certain boroughs demonstrate a **consistently higher risk profile**.

### **Demographic Trends**
- **Male victims aged 18-44** are the most affected demographic group.
- **Young males under 24** are particularly vulnerable to incidents.

**Recommendations**

### **Targeted Law Enforcement**
- **Deploy additional patrols** during peak hours (**10 PM - 2 AM**).
- Focus resources in **high-risk boroughs**, such as **Brooklyn**.

### **Community Engagement**
- **Develop prevention programs** targeting **young males under 24**.
- Collaborate with **community leaders** to promote safety initiatives.

### **Policy Adjustments**
- **Analyze seasonal trends** to optimize resource allocation during high-risk periods.
- **Increase investment in public safety infrastructure** in vulnerable boroughs.

```{r}


